---
title: "README"
output: github_document
bibliography: vignettes/bibfile.bib
---

```{r, setup, echo = FALSE, message=FALSE, warning=FALSE}
# load packages
library(tci)
library(knitr)
library(gridExtra)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align="center")
```

# tci

The `tci` package implements target-controlled infusion (TCI) algorithms to calculate infusion schedules for compartmental pharmacokinetic (PK) and pharmacokinetic-pharmacodynamic (PK-PD) models. Using these infusion schedules, functions in `tci` can be used to simulate PK or PK-PD responses with or without model misspecification. Generated responses can be used to simulate closed-loop control by implementing Bayesian updates to the PK or PK-PD model based on the "observed" (i.e., simulated) data. 

Closed-form solutions are provided for one, two, or three compartment mammillary models (i.e., all peripheral compartments are joined to a central compartment), as well as a three-compartment model with an adjoining effect-site. Alternative models, potentially based on ordinary differential equations (ODE), can be specified using other R packages and adapted for use with `tci` functions (see the "Custom PK models and algorithms" vignette). 

`tci` implements both plasma-and effect-site targeting TCI algorithms based on work by @Jacobs1990 and @Shafer1992, respectively. Users can implement alternative user-defined TCI algorithms, however. Again, see the "Custom PK models and algorithms" vignette for further details.

Several population PK models commonly used for TCI are implemented in `tci`. These include the Marsh, Schnider, and Eleveld models for propofol, and the Minto, Kim, and Eleveld models for remifentanil. Use of these models is illustrated in the "Population PK models" vignette.

 
## Installation

The `tci` package can be installed from [CRAN](https://cran.rstudio.com/web/packages/tci/index.html) using the command. 

```{r, eval = FALSE}
install.packages("tci")
```

The most recent version can be downloaded from GitHub using the `devtools` package and loaded as follows. 

```{r, eval = FALSE}
devtools::install_github("jarretrt/tci")
```

```{r, libraries}
library(tci)
library(ggplot2) # for plotting
```


## Examples

Equations implementing 1-, 2-, 3-compartment and 3-compartment-effect structural PK models are included in the `tci` package. The function `pkmod` will automatically infer the correct structure based on the parameter names. 

```{r}
# 3-compartment model with effect site
(mod3ecpt <- pkmod(pars_pk = c(cl = 10, q2 = 2, q3 =20, v = 15, v2 = 30, v3 = 50, ke0 = 1.2)))
```

Acceptable parameter names can be viewed by calling `list_parnms()`. Less-commonly used parameters, such as clearance from a peripheral compartment, are also permissible. 

```{r}
# acceptable parameter names
list_parnms()
```

Elements of `pkmod` objects can be modified through an `update.pkmod` method. Perhaps most usefully, this allows for partial modifications to PK-PD parameters. For example, the effect-site equilibrium constant can be easily updated.

```{r}
update(mod3ecpt, pars_pk = c(ke0 = 0.9), init = c(1,0.2,0.3,1))
```

Most functions in the `tci` package pass additional arguments to `update.pkmod` allowing for easy modification of `pkmod` objects as needed.


## TCI Infusion schedules

TCI algorithms are implemented using the function `tci_inf` (manual infusions are implemented by `inf_manual`). The user supplies a set of targets, times at which the target is set, and a `pkmod` object. The TCI algorithm (defaults to `type = "plasma"`) is iteratively applied to calculate infusion rates required to reach each target in turn. By default, infusion rates are updated in increments of 1/6, corresponding to every 10-second intervals if infusions rate units are in amount per minute. Infusion rates themselves must have the same units as the PK elimination parameters. If elimination rates are in different units, such as hours, then the TCI update frequency should be modified by the argument `dtm`.

```{r}
# effect-site targeting for three-compartment effect site model
inf_3ecpt <- inf_tci(target_vals = c(2,3,4,4), target_tms = c(0,2,3,10), 
                     pkmod = mod3ecpt, type = "effect")
head(inf_3ecpt)
```

## Population PK models

Population PK models are implemented by the function `poppkmod`. The user must supply a data frame with the set of covariates (e.g., weight, age) required by the model. Several published population PK models are currently implemented in `tci`. For propofol, these include the Marsh, Schnider, and Eleveld models. For remifentanil, they include the Minto, Kim, and Eleveld models. See `?poppkmod` or the population PK model vignette for details. `list_pkmods()` will list available population PK models and covariates required by each. 

```{r}
# data frame of patient covariates
data <- data.frame(ID = 1:5, AGE = seq(20,60,by=10), 
                   TBW = seq(60,80,by=5), HGT = seq(150,190,by=10), 
                   MALE = c(TRUE,TRUE,FALSE,FALSE,FALSE))
# Eleveld population PK model for propofol
pkpd_elvd <- poppkmod(data = data, drug = "ppf", model = "eleveld")
```

As with the `pkmod` class, `poppkmod` objects can be used by `inf_tci` and have `predict` and `simulate` methods to predict and simulate PK-PD responses, respectively.

PK-PD parameter values can be drawn at random from the inter-/intra-individual variability distribution, as described by the `pkmod` Omega matrix, by either 1) setting the argument `sample = TRUE` when calling `poppkmod`, or 2) by using the function `sample_iiv`. 

```{r}
set.seed(1)
pkpd_elvd_iiv <- sample_iiv(pkpd_elvd)
```


## Simulations

Simulations are best implemented through the function `simulate_tci`, which allows for model misspecification as well as Bayesian updates to model parameters based on previously observed data (i.e., "closed-loop" control). `simulate_tci` can be used for both `pkmod` or `poppkmod` classes.  Required arguments to `simulate_tci` are 1) a prior PK model (`pkmod_prior`) that is used to calculate infusion rates and may be updated throughout the simulation if update times are provided, 2) a true PK model (`pkmod_true`) that is used to simulate observations, 3) TCI target values, 4) TCI target times, and 5) times to simulate observations. 

```{r}
# TCI target values (PD response)
target_vals <- c(75,60,50,50)
# values are in terms of minutes. 1/6 = 10 seconds
# TCI target times
target_tms <- c(0,3,6,10)
# observation times 
obs_tms <- seq(1/6,10,1/6)

# simulate without updates ("open-loop")
sim_ol <- simulate_tci(pkmod_prior = pkpd_elvd, pkmod_true = pkpd_elvd_iiv, 
             target_vals, target_tms, obs_tms, type = "effect", seed = 1)
```

`simulate_tci` returns an object with class `sim_tci` that can be plotted using the `ggplot2` library.

```{r}
plot(sim_ol)
```

Modifications can be made to the plot to show a subset of responses, concentrations instead of PD response values, infusion rates, and simulated data.

```{r}
plot(sim_ol, yvar = "c4", id = c(1,3,5), show_inf = TRUE, wrap_id = TRUE)
```

Closed-loop simulations can be implemented by specifying a set of update times. We illustrate this with updates each minute and a processing delay of 20 seconds. 

```{r}
sim_cl <- simulate_tci(pkmod_prior = pkpd_elvd, pkmod_true = pkpd_elvd_iiv, 
             target_vals, target_tms, obs_tms, update_tms = 1:10, delay = 1/3,
               type = "effect", seed = 1)
```

Since `plot.sim_tci` returns a `ggplot2` object, it is easy to modify aspects such as titles and axis labels using `ggplot2` functions. 

```{r}
plot(sim_cl) + 
  xlab("Minutes") + 
  ylab("Bispectral Index") + 
  ggtitle("Closed-loop simulation of Eleveld propofol model", 
          subtitle = "Minute updates, processing delay of 20 seconds")
```



