---
title: "README"
output: github_document
bibliography: bibfile.bib
---

```{r, setup, echo = FALSE, message=FALSE, warning=FALSE}
# load packages
library(tci)
library(knitr)
library(gridExtra)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align="center")
```

# tci

The `tci` package can be used to apply target-controlled infusion (TCI) algorithms to 1-, 2-, and 3-compartment pharmacokinetic (PK) models with intravenous drug administration. Three-compartment PK models with an effect-site compartment are also supported. TCI algorithms incorporated in the package permit plasma or effect-site targeting and incorporation of closed-form pharmacodynamic (PD) models. Alternate user-defined PK models, population PK models, or TCI algorithms can additionally be specified. 

## Installation

The `tci` package can be installed from GitHub using the `devtools` package and loaded as follows. 

```{r, eval = FALSE}
devtools::install_github("jarretrt/tci")
library(tci)
```


## Examples

### PK Models

PK models can be evaluated by the `predict.pkmod` method in combination with a dosing schedule. Dosing schedules should indicate infusion starting times and rates, with the infusion termination time specified with rate 0. 


```{r, dose-object}
# e.g. infusion rates of 100 mg/hr for 30 sec intervals at 0, 4 minutes.
dose <- create_intvl(
  as.matrix(cbind(time = c(0.5,4,4.5,10), 
                  infrt = c(100,0,100,0)))
)
dose

# model parameters 
pars_3cpt <- c(k10=1.5,k12=0.15,k21=0.09,k13=0.8,
               k31=0.8,v1=10,v2=15,v3=100,ke0=1)

# predict concentrations of a three-compartment model with effect-site at
# times 1, 2, 8 minutes
predict(pkmod3cptm, 
        pars = pars_3cpt,
        inf = dose, 
        tms = c(1,2,8))

# plot concentrations
plot(pkmod3cptm, inf = dose, pars = pars_3cpt,
     title = "Concentrations for a 3 compartment model with an effect site")
```

Closed-form PK model solutions for 1-, 2-, and 3-compartment models are provided by the `tci` package based on solutions and code in @Abuhelwa2015. Custom user-defined PK models, potentially using ODE solvers available in other packages, can be specified and used. See the vignette on user-defined PK functions for an example.



### TCI algorithms

TCI algorithms are iteratively applied through the function 'tci'. Times and target concentrations are passed as arguments along with a PK model, PK model parameters, and (optional) initial concentrations. Times and targets should include the time/concentration associated with the end of the infusion schedule. Plasma or effect-site targeting is specified through the "tci_alg" argument, or a custom TCI algorithm is provided through the argument "tci_custom." 

```{r, tci-algs}
# target concentrations of 2 for 0-5 minutes and 3 for 5-10 minutes.
tci_times <- c(0,5,10,10)
tci_targets <- c(2,3,3,3)

# plasma-targeting
inf_3cpt_plasma <- tci(Ct = tci_targets, 
                       tms = tci_times, 
                       pkmod = pkmod3cptm, 
                       pars = pars_3cpt, 
                       tci_alg = "plasma")

# infusions for effect-site targeting
inf_3cpt_effect <- tci(Ct = tci_targets, 
                       tms = tci_times, 
                       pkmod = pkmod3cptm, 
                       pars = pars_3cpt, 
                       tci_alg = "effect", 
                       cptol = 0.2)

head(inf_3cpt_effect)
```

Infusion schedules are provided in the output of 'tci' along with predicted concentrations and can be plotted with the "plot.tciinf" method.

```{r, plot-tci, fig.height=10}
ptci_2ug_plasma <- plot(inf_3cpt_plasma, 
                        title = "Plasma targeting for three-compartment model",
                        display = FALSE)

ptci_2ug_effect <- plot(inf_3cpt_effect, 
                        title = "Effect-site targeting for three-compartment model",
                        display = FALSE)

grid.arrange(ptci_2ug_plasma, ptci_2ug_effect)
```


### Population PK models

Functions implementing the Marsh, Schnider, and Eleveld population PK models for propofol (PK-PD for Eleveld model) are currently provided and can be used to calculate PK and PK-PD parameters at patient covariates or to simulate parameters for new patients given inter-patient parameter variability. 

```{r, poppk}
patient_dat <- data.frame(AGE  = c(20,40,65),
                          TBM  = c(50,70,90),
                          HGT  = c(150,170,200),
                          MALE = c(TRUE,FALSE,TRUE))

# evaluate at covariate values and return clearance parameters
patient_pk <- schnider_poppk(patient_dat, rand = FALSE, rate = FALSE)
patient_pk

# evaluate TCI for patient 1
tci_patient1 <- tci(Ct = tci_targets, 
                    tms = tci_times, 
                    pkmod = pkmod3cptm, 
                    pars = patient_pk[1,], 
                    tci_alg = "effect")
head(tci_patient1)
```

`tci` functions are compatible with other user-specified population PK models provided that they return parameters with either elimination rate-constants named "k10", "k12", etc. (capitalization optional), or as clearance parameters "CL", "Q2", etc.


## References






