---
title: "Simulation from a user-specified, ODE-based PK model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation from a user-specified, ODE-based PK model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibfile.bib  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### User-defined PK functions using the tci package

The `tci` package provides closed-form solutions for 1-, 2-, 3-, and 3-compartment with effect-site PK models based on solutions described and code provided in @Abuhelwa2015. It does not, however, include an ordinary differential equation (ODE) solver as many other R packages do, including [mrgsolve](https://mrgsolve.github.io/), [PKPDsim](http://pkpdsim.ronkeizer.com/), and [RxODE](https://github.com/nlmixrdevelopment/RxODE). `tci` package functions can nonetheless be applied to models from these packages through the creation of user-defined PK functions.
 
User-defined PK functions should be created to return concentrations at a vector of time points under a continuous infusion and should be assigned S3 class "pkmod." To illustrate, we create an example two-compartment model based on ODEs is illustrated using the `mrgsolve` R package[@mrgsolve]. Using `mrgsolve` syntax, a model can be specified as a character string and compiled using the `mcode` function.

```{r, mrgsolve-example, warning=FALSE}
library(tci)
library(mrgsolve)

# define two-compartment model based on ODEs
code <- '
$PARAM CL=10, V=10, Q = 10, V2=10
$CMT A1 A2
$ODE
dxdt_A1 = -(CL/V)*A1 - (Q/V)*A1 + (Q/V2)*A2;
dxdt_A2 = -(Q/V2)*A2 + (Q/V)*A1;
'
mod2cpt <- mcode("example", code)
```

Initial parameter values are required by `mrgsolve` but can be updated as needed using the `mrgsolve::update` function. We now need to write a wrapper that takes as arguments a time at which to evaluate the PK model (`tm`), an infusion rate (`kR`), PK model parameters (`pars`), initial concentrations (`init`), and an infusion starting time (`inittm`). The starting time is used by functions in `tci` that iteratively calculate concentrations from a series of infusions. The function should return a matrix with concentrations for each compartment in the rows and evaluation times as columns.

```{r}
# wrapper function to evaluate pk2 in format used by tci
pkmod2cpt_ode <- function(tm, kR, pars, 
                          pkm = pk2, 
                          init = c(0,0), 
                          inittm = 0){

  # begin times at 0 and end at last time evaluated
  tm <- tm - inittm
  end_inf <- max(tm)
  
  # name parameter/initial concentrations
  names(pars) <- toupper(names(pars))
  names(init) <- c("A1","A2")
  
  # pass parameters as list
  pars <- sapply(pars, as.list)
  vols <- unlist(pars[c("V","V2")])
  
  # update parameters and initial values (as amounts)
  mod2cpt <- update(mod2cpt, param = pars, init = init*vols)
  
  # dosing regimen - mrgsolve function in terms of amount infused
  event <- ev(amt =  kR*end_inf, time = 0, tinf = end_inf)
  
  # simulate responses (skip tm=0 unless specified)
  dat <- mrgsim_q(x = mod2cpt, # pk model
                  data = event, # dosing event
                  stime = tm) # evaluation times
  
  # skip tm=0 unless specified in tm
  dat <- dat@data[-1,]
  
  # return concentrations with compartments in rows and times in columns
  cons <- t(dat[,c("A1","A2")]) / vols
  
  rownames(cons) <- colnames(cons) <- NULL
  return(cons)
}
class(pkmod2cpt_ode) <- "pkmod"
```

For use with the `tci` package, the user-defined PK function should have class "pkmod". We can now assign parameter values and predict from this model as we would any other. 

```{r, fig.height= 6, fig.width = 6}
# create dose
dose <- create_intvl(
  as.matrix(cbind(time = c(0.5,4,4.5,10), 
                  infrt = c(100,0,100,0)))
)

# model parameters
pars_2cpt <- c(CL = 15, V = 10, Q = 10, V2 = 20)

# predict concentrations
predict(pkmod2cpt_ode, 
        pars = pars_2cpt, 
        inf = dose, 
        tms = c(1,2,3))

# plot concentrations
plot(pkmod2cpt_ode, 
     pars = pars_2cpt, 
     inf = dose, 
     title = "Concentrations for a user-defined 2-compartment model")
```


We can also apply TCI algorithms to calculate infusion schedules required to reach designated targets with the user-defined model. 

```{r, user-defined-tci, fig.height= 6, fig.width = 6}
# target concentration = 1 for times 0 - 5
# target concentration = 2 for times 5 - 10
tci_times <- c(0,5,10)
tci_targets <- c(1,2,2)

# plasma-targeting
inf_3cpt_plasma <- tci(Ct = tci_targets, 
                       tms = tci_times, 
                       pkmod = pkmod2cpt_ode, 
                       pars = pars_2cpt, 
                       tci_alg = "plasma")
# print infusion schedule
head(inf_3cpt_plasma)

# plot infusion schedule
plot(inf_3cpt_plasma)
```


### References




